Parameter Recovery Validation

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): standardize on `t^*_{on,i}`
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

$$\alpha^*_i(t^*) := \begin{cases}
1.0 & \text{if } t^* < t^*_{0,\text{on}i}, \\
R_{\text{on},i} & \text{if } t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i, \\
1.0 & \text{if } t^* \geq t^*_{0,\text{on}i} + \delta^*_i
\end{cases}$$

@@

- $\alpha^*_{\text{off}i} = 1.0$ is the fixed reference transcription rate
- $R_{\text{on},i} = \alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} = \alpha^*_{\text{on}i}$ is the fold-change during activation (inferred parameter)
- $t^*_{0,\text{on}i}$ is the dimensionless activation onset time (can be negative for pre-activation scenarios)
- $\delta^*_i$ is the dimensionless activation duration

@@

When $t^*_{0,\text{on}i} < 0$:

@@

#### **Phase 1: Off State** ($t^* < t^*_{0,\text{on}i}$)

@@

**Note**: This phase may not be observed if $t^*_{0,\text{on}i} < 0$ (pre-activation scenario).

#### **Phase 2: On State** ($t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i$)

Let $\tau_{\text{on}} = t^* - t^*_{0,\text{on}i}$ be the time since activation onset. The system transitions from the off-state steady state toward the on-state steady state.

@@

#### **Phase 3: Return to Off State** ($t^* \geq t^*_{0,\text{on}i} + \delta^*_i$)

Let $\tau_{\text{off}} = t^* - (t^*_{0,\text{on}i} + \delta^*_i)$ be the time since returning to the off state. The system transitions from the end-of-phase-2 state back toward the off-state steady state.

@@

- $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ (early activation within process duration)
- $\delta^*_i < 0.4 \mathbb{E}[T_M]$ (short activation duration relative to process)
- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)

@@

- $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ (early activation within process duration)
- $\delta^*_i > 0.5 \mathbb{E}[T_M]$ (long activation duration relative to process)
- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)

@@

**General principle:**
\begin{align}
t^*_{on,i} &\sim \text{Normal}(f_{on} \times \mathbb{E}[T_M], \sigma_{on}^2) \\
\delta^*_i &\sim \text{LogNormal}(\log(f_{\delta} \times \mathbb{E}[T_M]), \sigma_{\delta}^2)
\end{align}

@@

**Concrete examples for different experimental contexts:**
- **Conservative (Gamma(5,1))**: $\mathbb{E}[T_M] = 5$ → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
- **Skeptical (Gamma(1,1))**: $\mathbb{E}[T_M] = 1$ → $t^*_{on,i} \sim \text{Normal}(0.3, \sigma_{on}^2)$
- **Permissive (Gamma(1,0.2))**: $\mathbb{E}[T_M] = 5$ → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$

@@

The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while using independent temporal parameterization to eliminate scaling symmetry. For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of independent temporal gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$. Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.

@@

- **Pre-activation**: $t^*_{on,i} < 0.0$ and $R_{\text{on},i} > 2.0$
- **Transient**: $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$, $\delta^*_i < 0.4 \mathbb{E}[T_M]$, and $R_{\text{on},i} > 2.0$
- **Sustained**: $0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$, $\delta^*_i > 0.5 \mathbb{E}[T_M]$, and $R_{\text{on},i} > 2.0$

where $\mathbb{E}[T_M] = \alpha_T/\beta_T$ for the Gamma prior, so these thresholds adapt to the chosen hyperparameters.

@@

- **$t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around $0.3 \times \mathbb{E}[T_M]$ to expect activation at 30% of process duration. **Negative values** enable pre-activation scenarios where genes were already activated before observation began.

- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around $0.4 \times \mathbb{E}[T_M]$ to expect 40% duration activation. This enables both transient (short duration) and sustained (long duration) activation patterns.

@@

- **Pre-activation** ($t^*_{on,i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
- **Transient** ($0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ and $\delta^*_i < 0.4 \mathbb{E}[T_M]$): Balanced probability for observable complete cycles
- **Sustained** ($0 < t^*_{on,i} < 0.3 \mathbb{E}[T_M]$ and $\delta^*_i > 0.5 \mathbb{E}[T_M]$): Balanced probability for long-duration activation

>>>>

$$\alpha^*_i(t^*) := \begin{cases}
1.0 & \text{if } t^* < t^*_{on,i}, \\
R_{\text{on},i} & \text{if } t^*_{on,i} \leq t^* < t^*_{on,i} + \delta^*_i, \\
1.0 & \text{if } t^* \geq t^*_{on,i} + \delta^*_i
\end{cases}$$

@@

- $\alpha^*_{\text{off}i} = 1.0$ is the fixed reference transcription rate
- $R_{\text{on},i} = \alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} = \alpha^*_{\text{on}i}$ is the fold-change during activation (inferred parameter)
- $t^*_{on,i}$ is the dimensionless activation onset time (can be negative for pre-activation scenarios)
- $\delta^*_i$ is the dimensionless activation duration

@@

@@

#### **Phase 1: Off State** ($t^* < t^*_{on,i}$)

@@

**Note**: This phase may not be observed if $t^*_{on,i} < 0$ (pre-activation scenario).

#### **Phase 2: On State** ($t^*_{on,i} \leq t^* < t^*_{on,i} + \delta^*_i$)

Let $\tau_{\text{on}} = t^* - t^*_{on,i}$ be the time since activation onset. The system transitions from the off-state steady state toward the on-state steady state.

@@

#### **Phase 3: Return to Off State** ($t^* \geq t^*_{on,i} + \delta^*_i$)

Let $\tau_{\text{off}} = t^* - (t^*_{on,i} + \delta^*_i)$ be the time since returning to the off state. The system transitions from the end-of-phase-2 state back toward the off-state steady state.

@@

- $0 < t^*_{on,i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
- $\delta^*_i < 2.0$ (short activation duration relative to process, assuming conservative prior Gamma(5,1))
- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)

@@

- $0 < t^*_{on,i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
- $\delta^*_i > 2.5$ (long activation duration relative to process, assuming conservative prior Gamma(5,1))
- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)

@@

**General principle:**
\begin{align}
t^*_{on,i} &\sim \text{Normal}(f_{on} \times \text{E}[T_{\max}], \sigma_{on}^2) \\
\delta^*_i &\sim \text{LogNormal}(\log(f_{\delta} \times \text{E}[T_{\max}]), \sigma_{\delta}^2)
\end{align}

@@

**Concrete examples for different experimental contexts:**
- **Conservative (Gamma(5,1))**: E[$T_{\max}$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$
- **Skeptical (Gamma(1,1))**: E[$T_{\max}$] = 1 → $t^*_{on,i} \sim \text{Normal}(0.3, \sigma_{on}^2)$
- **Permissive (Gamma(1,0.2))**: E[$T_{\max}$] = 5 → $t^*_{on,i} \sim \text{Normal}(1.5, \sigma_{on}^2)$

@@

The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while using independent temporal parameterization to eliminate scaling symmetry. For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of independent temporal gene parameters $(t^*_{on,i}, \delta^*_i)$. Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.

@@

- **Pre-activation**: $t^*_{on,i} < 0.0$ and $R_{\text{on},i} > 2.0$
- **Transient**: $0 < t^*_{on,i} < 1.5$, $\delta^*_i < 2.0$, and $R_{\text{on},i} > 2.0$ (assuming conservative prior Gamma(5,1))
- **Sustained**: $0 < t^*_{on,i} < 1.5$, $\delta^*_i > 2.5$, and $R_{\text{on},i} > 2.0$ (assuming conservative prior Gamma(5,1))

@@

- **$t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around 1.5 to expect activation at 30% of process duration (assuming conservative prior Gamma(5,1) with mean 5). **Negative values** enable pre-activation scenarios where genes were already activated before observation began.

- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around 2.0 to expect 40% duration activation (assuming conservative prior Gamma(5,1) with mean 5). This enables both transient (short duration) and sustained (long duration) activation patterns.

@@

- **Pre-activation** ($t^*_{on,i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
- **Transient** ($0 < t^*_{on,i} < 1.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles (assuming conservative prior Gamma(5,1))
- **Sustained** ($0 < t^*_{on,i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation (assuming conservative prior Gamma(5,1))
